Allee Effect and Cultural Evolution

### 2015-07-07 manuscript/draft.tex
small fixes

\begin{abstract}
Several social learning strategies guide the selection of cultural variant to imitate through an evaluation of the expected benefits associated with each observed alternatives. The assumption of the social learner is that the payoff received by the transmitter while expressing a given cultural variant is a proxy for  inferring the qualitative features of the variant itself. Here we explore the theoretical implications when: 1) the payoff attributed to a given cultural variant directly affects the reproductive success of its bearer;  and 2) payoff is density dependent, that is its expected amount is a function of the number of individuals possessing the same variant. We consider situations where such density dependence is simultaneously direct and inverse, with minimum and maximum critical thresholds bounding the range of positive payoff, with a peak value at midpoint. This non-linear relationship, known as Allee effect, portrays a large number of behavioural traits that are enhanced by cooperation or mutual facilitation, but are somehow constrained by some form of carrying capacity.  We explore the evolutionary dynamics of a single trait-two variants model, each exhibiting a different Allee effect. We used a modified Lotka-Volterra model of payoff-biased social learning, and show that different degrees of reliance on social learning can exhibit a variety of equilibria, including episodes of reversion, stable co-existence, and fixation.

@@

Several behavioural traits are expected to show such a density dependence. For example, behaviour linked to the exploitation of limited resources will exhibit a decline in payoff when density exceeds a critical threshold (i.e carrying capacity). In this case a new optimal variants are likely to spread rapidly when the they are rare, but as the number of individuals possessing these approach their respective carrying capacities, their beneficial effects will start to decline, the payoff  decrease, and eventually rarer suboptimal variants might be preferred (see Lake and Crema \citeyear{lake_and_crema_2012} for an extensive analysis).

@@

For the rest of the paper we will consider only scenarios where the node $x=K_x$ is stable. In this case the initial value of $x(t=0)$ will determine the ultimate equilibria of the system: extinction, if $x(t=0)<A_x$; unstable equilibrium, if $x(t=0)=A_x$; and stable equilibrium, if $x(t=0)>A_x$. Thus in a 2-trait model, the coexistence of both variants is guaranteed only when the initial density of both traits are above their respective critical densities, otherwise we should expect that either (i.e. $a(t=0) \geq A_a$ and $b(t=0)<A_b$ or $a(t=0)<A_a$ and $b(t=0) \geq A_b$) or both (i.e. when $a(t=0)<A_a$ and $b(t=0)<A_b$) variants are ultimately extinct (see fig. \ref{fig:NoTransmissionBasin}).

@@

First, the basins of attraction are transformed, and hence the system equilibrium at $z=0$ and $z>0$ can be different for the same initial conditions. This is particularly the case for combinations of $a$ and $b$ that are closer to the boundaries between two or more basins.  Figure \ref{fig:TransmissionBasin} shows an example of this. When social learning is not present (fig.~\ref{fig:TransmissionBasin}a), the initial conditions $i$, $j$, and $k$ result into a stable equilibrium at $a=K_a$ and $b=K_b$, whilst the starting point $l$ leads to the extinction of $b$. When $z=0.05$ (fig.~\ref{fig:TransmissionBasin}b), the basins of attractions are modified, resulting into different trajectories and in some cases different equilibria for the same initial conditions. Thus $i$ and $k$ are no longer leading to the coexistence of both traits but to the demise of one or the other, whilst $l$ now ensures the coexistence of both variants.

@@

Second, when the magnitude of social learning is high, the system is subject to effects similar to those observed with an increase in the per capita growth rate $r$ (cf. fig. \ref{fig:bifurcationDiagram}), that is a decrease in the stability of the fixed points. Figure \ref{fig:BifurcationSocialLearning} shows the bifurcation diagram for the population size of $a$ as a function of $z$. Here initial the initial effects of $z$ does not alter the equilibrium point ($K_a=0.6$), but once $z>0.5$ we observe a sudden transition to unstable fluctuations, with an increasing range of different values for $a$. %We can keep more or less the same description showing instead \ref{fig:TSsocialLearning}

The basin plot (fig.\ref{fig:basinunstable}) suggests that this regime is prone to occur when both variants are initialised with high densities in respect to their critical densities. In these conditions the total density is too high to ensure a stable coexistence, and the regulatory effect of negative fitness is minimised by the fast rates of cultural transmission (i.e. individual shift strategy before experiencing a population decline resulting from density being over $K$).

\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{./figures/figure4.jpg}
  \caption{}
    \label{fig:BifurcationSocialLearning}
\end{figure}

@@

The results detailed in previous section indicates that, within the positive payoff region, fixation to one strategy or another can occur only for a comparatively narrow set of initial conditions close to the boundaries between different basins of attractions. For all other conditions we expect to observe either a stable coexistence or unstable fluctuations depending on how much individuals are prone to social learning. The results thus implies that if two variants have the similar payoff values, there is no intrinsic advantage of one strategy over the other. Thus the variant with higher critical thresholds $A$ and $K$ do not necessarily predominate over a wider range of initial population sizes, and instead its full adoption by the entire population $N$ is ensure only when its initial density is comparatively high (e.g. above or equal its maximum payoff) and when the alternative strategy is close to its critical threshold $A$. This also implies that if the variant with higher $A$ and $K$ is novel, some mechanism (e.g. an alternative form of social learning) must ensure that its density approaches its highest payoff, and at the same, the incumbent strategy should exhibit a comparatively low density. Alternatively, the novel variant should either exhibit a higher payoff (thus has some intrinsic advantage) or some form of competition between the two variant should be considered. Here we briefly explore the latter scenario, by relaxing the assumption of independence in the computation of the payoff of each variant. For instance several subsistence strategies do not have fully independent resource pools, and hence the success of a given variant might be detrimental to the payoff of the other irregardless of the effects provoked by cultural transmission. We can revise our core model by including new terms portraying the competition effect between the two resources:

@@

\begin{figure*}[h!]
  \centering
      \includegraphics[width=\textwidth]{./figures/figure7}
  \caption{The impact of competition on the basins of attraction for the different equilibria. The Parameters used are $z=0.05$ (left) and $z=0.2$ (right) for social learning, and $c=-0.1$ (top) and $c=0.1$ (bottom) for competition.}
    \label{fig:competition}
\end{figure*}

@@

Our analysis examined the impact of payoff-based social learning strategies in the evolutionary dynamics of cultural traits with direct and inverse density dependence. The results showed that social learning can alter the basins of attraction of stable nodes, leading to different dynamics and equilibria from the very same initial conditions. If there is no reliance on social learning, the critical density separates the ultimate equilibrium of the system: if the initial condition is below this threshold the variant is destined to go extinct, whilst if it is above this value it will eventually reach carrying capacity. When social learning is present this boundary shifts, variants can survive even if its initial condition is below critical density and equally can reach extinction even if above this threshold. This scenario is further enhanced when the two variants are not independent but are in direct competition. In this case, social learning can further enhance the advantage of one variant over the other.

When the reliance on social learning is comparatively high, and  he two variants have a substantial overlap in their range of positive payoff densities (i.e. they share similar critical densities and carrying capacities), the system does not always reach an stable equilibrium, but can instead exhibit cyclical fluctuations between the two variants. In other words the consequences of social learning is not absorbed by the system, and instead a negative feedback process leads to continuous switch from one variant to the other.

@@

The implications of our model can provide some insights in the evolutionary history of cultural traits exhibiting density dependence. Perhaps the most notable example is the spread of keystone innovations in subsistence strategies which led to the adoption of variants enabling larger carrying capacity, often at the cost of higher critical densities. The spread of behavioural traits such as the know-how of the domestication process~\citep{barker2006} the use of new subsistence tools~\citep{petraglia_population_2009} and processing techniques~\citep{molleson1993}, as well as the inclusion/exclusion and extensification/intensification of different prey species are just some of the most notable examples. In many of these examples the initial spread of the innovation generated a positive feedback process, either through enhanced possibility of direct cooperation, or some form of mutual facilitation often indirectly emerging from positive niche construction. At the same time, ethnographic and archaeological evidence show how many instances these critical transitions where not smooth, often leading to the coexistence of alternative strategies as well as episodes of a return to a previously abandoned cultural traits. Several ethnographic examples showcase small-scale societies between alternating between different subsistence modes over a comparatively brief period of time \citep{layton1991,mace_1993,oota_et_al_2005},
 while increasing archaeological evidence suggest that the transition from hunting-gathering to farming was not unidirectional, but with instances of temporary reversions  \citep{rowley2001,stevens_fuller_2012} %ideally we need a couple of more citations here?
While in most cases these are  anecdotal evidence rather than testable empirical cases, the insights of our models suggests that different levels of reliance on social learning could alone portray some of the evolutionary trajectories observed in these systems.

>>>>

\begin{abstract}
Several social learning strategies guide the selection of cultural variant to imitate through an evaluation of the expected benefits associated with each observed alternatives. The assumption of the social learner is that the payoff received by the transmitter while expressing a given cultural variant is a proxy for  inferring the qualitative features of the variant itself. Here we explore the theoretical implications when: 1) the payoff attributed to a given cultural variant directly affects the reproductive success of its bearer;  and 2) payoff is density dependent, that is its expected amount is a function of the number of individuals possessing the same variant. We consider situations where such density dependence is simultaneously direct and inverse, with minimum and maximum critical thresholds bounding the range of positive payoff, with a peak value at midpoint. This non-linear relationship, known as Allee effect, portrays a large number of behavioural traits that are enhanced by cooperation or mutual facilitation, but are somehow constrained by some form of carrying capacity.  We explore the evolutionary dynamics of a single trait-two variants model, each exhibiting a different Allee effect. We use a modified Lotka-Volterra model of payoff-biased social learning, and show that different degrees of reliance on social learning can exhibit a variety of equilibria, including episodes of reversion, stable co-existence, and fixation.

@@

Several behavioural traits are expected to show such a density dependence. For example, behaviour linked to the exploitation of limited resources will exhibit a decline in payoff when density exceeds a critical threshold (i.e carrying capacity). In this case new optimal variants are likely to spread rapidly when they are rare, but as the number of individuals possessing them approach their respective carrying capacities, their beneficial effects will start to decline, the payoff will decrease, and eventually rarer suboptimal variants might be preferred (see Lake and Crema \citeyear{lake_and_crema_2012} for an extensive analysis).

@@

For the rest of the paper we will consider only scenarios where the node $x=K_x$ is stable. In this case the initial value of $x_{t=0}$ will determine the ultimate equilibria of the system: extinction, if $x_{t=0}<A_x$; unstable equilibrium, if $x_{t=0}=A_x$; and stable equilibrium, if $x_{t=0}>A_x$. Thus in a 2-trait model, the coexistence of both variants is guaranteed only when the initial density of both traits are above their respective critical densities, otherwise we should expect that either (i.e. $a_{t=0} \geq A_a$ and $b_{t=0}<A_b$ or $a_{t=0}<A_a$ and $b_{t=0} \geq A_b$) or both (i.e. when $a_{t=0}<A_a$ and $b_{t=0}<A_b$) variants are ultimately extinct (see fig. \ref{fig:NoTransmissionBasin}).

@@

First, the basins of attraction are transformed, and hence the system equilibrium at $z=0$ and $z>0$ can be different for the same initial conditions. This is particularly the case for combinations of $a$ and $b$ that are closer to the boundaries between two or more basins.  Figure \ref{fig:TransmissionBasin} shows an example of this. When social learning is not present (fig.~\ref{fig:TransmissionBasin} left), the initial conditions $i$, $j$, and $k$ result into a stable equilibrium at $a=K_a$ and $b=K_b$, whilst the starting point $l$ leads to the extinction of $b$. When $z=0.05$ (fig.~\ref{fig:TransmissionBasin} right), the basins of attractions are modified, resulting into different trajectories and in some cases different equilibria for the same initial conditions. Thus $i$ and $k$ are no longer leading to the coexistence of both traits but to the demise of one or the other, whilst $l$ now ensures the coexistence of both variants.

@@

Second, when the magnitude of social learning is high, the system is subject to effects similar to those observed with an increase in the per capita growth rate $r$ (cf. fig. \ref{fig:bifurcationDiagram}), that is a decrease in the stability of the fixed points. Figure \ref{fig:BifurcationSocialLearning} shows the bifurcation diagram for the population size of $a$ as a function of $z$. Here the initial effects of $z$ do not alter the equilibrium point ($K_a=0.6$), but once $z>0.5$ we observe a sudden transition to unstable fluctuations, with an increasing range of different values for $a$. %We can keep more or less the same description showing instead \ref{fig:TSsocialLearning}

% TODO XRC I thought we were removing this figure
The basin plot (fig.\ref{fig:basinunstable}) suggests that this regime is prone to occur when both variants are initialised with high densities in respect to their critical densities. In these conditions the total density is too high to ensure a stable coexistence, and the regulatory effect of negative fitness is minimised by the fast rates of cultural transmission (i.e. individual shift strategy before experiencing a population decline resulting from density being over $K$).

% TODO XRC choose figure4 or figure4b
\begin{figure}
  \centering
      \includegraphics[width=0.5\textwidth]{./figures/figure4.jpg}
  \caption{}
    \label{fig:BifurcationSocialLearning}
\end{figure}

@@

The results detailed in previous section indicates that, within the positive payoff region, fixation to one strategy or another can occur only for a comparatively narrow set of initial conditions close to the boundaries between different basins of attractions. For all other conditions we expect to observe either a stable coexistence or unstable fluctuations depending on how much individuals are prone to social learning. The results thus implies that if two variants have similar payoff values, there is no intrinsic advantage of one strategy over the other. Thus the variant with higher critical thresholds $A$ and $K$ do not necessarily predominate over a wider range of initial population sizes, and instead its full adoption by the entire population $N$ is ensured only when its initial density is comparatively high (e.g. above or equal its maximum payoff) and when the alternative strategy is close to its critical threshold $A$. This also implies that if the variant with higher $A$ and $K$ is novel, some mechanism (e.g. an alternative form of social learning) must ensure that its density approaches its highest payoff, and at the same, the incumbent strategy should exhibit a comparatively low density. Alternatively, the novel variant should either exhibit a higher payoff (thus has some intrinsic advantage) or some form of competition between the two variant should be considered. Here we briefly explore the latter scenario, by relaxing the assumption of independence in the computation of the payoff of each variant. For instance several subsistence strategies do not have fully independent resource pools, and hence the success of a given variant might be detrimental to the payoff of the other irregardless of the effects provoked by cultural transmission. We can revise our core model by including new terms portraying the competition effect between the two resources:

@@

\begin{figure*}[h!]
  \centering
      \includegraphics[width=\textwidth]{./figures/figure7}
  \caption{The impact of competition on the basins of attraction for the different equilibria. The Parameters used are $z=0.05$ (up) and $z=0.2$ (bottom) for social learning, and $c=0$ (left), $c=0.1$ (center) and $c=0.2$ (right) for competition.}
    \label{fig:competition}
\end{figure*}

@@

Our analysis examined the impact of payoff-based social learning strategies in the evolutionary dynamics of cultural traits with direct and inverse density dependence. The results showed that social learning can alter the basins of attraction of stable nodes, leading to different dynamics and equilibria from the very same initial conditions. If there is no reliance on social learning, the critical density separates the ultimate equilibrium of the system: if the initial condition is below this threshold the variant is destined to go extinct, whilst if it is above this value will eventually reach carrying capacity. When social learning is present this boundary shifts, variants can survive even if its initial condition is below critical density and equally can reach extinction even if above this threshold. This scenario is further enhanced when the two variants are not independent but are in direct competition. In this case, social learning can further enhance the advantage of one variant over the other.

When the reliance on social learning is comparatively high, and the two variants have a substantial overlap in their range of positive payoff densities (i.e. they share similar critical densities and carrying capacities), the system does not always reach an stable equilibrium, but can instead exhibit cyclical fluctuations between the two variants. In other words the consequences of social learning is not absorbed by the system, and instead a negative feedback process leads to continuous switch from one variant to the other.

@@

The implications of our model can provide some insights in the evolutionary history of cultural traits exhibiting density dependence. Perhaps the most notable example is the spread of keystone innovations in subsistence strategies which led to the adoption of variants enabling larger carrying capacity, often at the cost of higher critical densities. The spread of behavioural traits such as the know-how of the domestication process~\citep{barker2006} the use of new subsistence tools~\citep{petraglia_population_2009} and processing techniques~\citep{molleson1993}, as well as the inclusion/exclusion and extensification/intensification of different prey species are just some of the most notable examples. In many of these examples the initial spread of the innovation generated a positive feedback process, either through enhanced possibility of direct cooperation, or some form of mutual facilitation often indirectly emerging from positive niche construction. At the same time, ethnographic and archaeological evidence show how many instances of critical transitions where not smooth, often leading to the coexistence of alternative strategies as well as episodes of a return to a previously abandoned cultural traits. Several ethnographic examples showcase small-scale societies alternating between different subsistence modes over a comparatively brief period of time \citep{layton1991,mace_1993,oota_et_al_2005}, while increasing archaeological evidence suggest that the transition from hunting-gathering to farming was not unidirectional, but with instances of temporary reversions \citep{rowley2001,stevens_fuller_2012} %ideally we need a couple of more citations here?
While in most cases these are anecdotal evidence rather than testable empirical cases, the insights of our model suggests that different levels of reliance on social learning could alone portray some of the evolutionary trajectories observed in these systems.
